A Large-scale Study of Polymorphism in R

### 2019-04-01 paper/main.tex
some type name macros

@@

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in \RLOC lines of R 
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by \PROGRAMMERS programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our result suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.
\end{abstract}

@@

	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt scalar/double} or {\tt vector/double} accordingly.
	This is also true for integer, complex, raw, logical, and character {\tt typeof}s (i.e., the primitive vectorized data types).
	A notable imperfection here is is that unit length vectors will appear as scalars, so later in our analysis we ensure that argument signatures containing both scalar/X and vector/X are collapsed down to contain only vector/X.

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat)) {
        stop(simpleError("...'dat' must be a vector!"))
      }
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a data.frame/(vector/double) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) {
        x <- t(x)
    }
    ...
\end{lstlisting}}\caption{Example of matrix/(vector/double) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}

@@

\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \tt scalar/double & \tt numeric & \tt \{\} \\
y & \tt vector/double & \tt numeric & \tt \{\} \\
retv & \tt list<double> & \tt list & \tt \{example: scalar/character\}  
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  vector/double, vector/integer & 7742 & 9.66 \\ 
  scalar/double, scalar/integer & 7201 & 8.99 \\ 
  matrix, vector/double & 5384 & 6.72 \\ 
  NULL, scalar/character & 2325 & 2.90 \\ 
  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
  data.frame, matrix & 2116 & 2.64 \\ 
  NULL, scalar/double & 2082 & 2.60 \\ 
  NULL, vector/double & 2069 & 2.58 \\ 
  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
  NULL, vector/character & 1380 & 1.72 \\
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the L1.}
\end{table}

We would like to make one important note before digesting the data in Table~\ref{tab:toppolyL1}.
We drew a distinction between scalars and vectors of the same type, and that distinction would be the single largest source of polymorphism in L1 had we not dealt with it appropriately.
We note that an argument which exhibits a signature including e.g. \{scalar/double, vector/double\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
We have already collapsed this distinction.

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Tue Mar 26 17:46:45 2019
\begin{table}[ht]
\label{tab:toppolyL1p}
\centering
\begin{tabular}{lrr}
  \hline
Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  matrix, vector/real & 6360 & 13.94 \\ 
  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
  vector/character, vector/real & 2774 & 6.08 \\ 
  scalar/character, scalar/real & 2552 & 5.59 \\ 
  data.frame, matrix & 2167 & 4.75 \\ 
  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\ 
  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\ 
  data.frame, vector/real & 1046 & 2.29 \\ 
  scalar/logical, scalar/real & 920 & 2.02 \\ 
  scalar/character, vector/real & 868 & 1.90 \\
   \hline
\end{tabular}
\caption{Top 10 argument signatures in L$1^{+}$.}
\end{table}

A few of the top polymorphic signatures stand out.
First, the matrix/(vector/real) signature is interesting: \AT{as we showed in our polymorphism examples section}, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
v <- c(1, 2, 3, 4)
ma <- matrix(data=v, nrow=4, ncol=1)
v[2] == m[2] == m[2, 1]

>>>>

% Macros for type names?
\newcommand{\reals}{$real$\xspace}
\newcommand{\realv}{$\overline{real}$\xspace}
\newcommand{\chars}{$char$\xspace}
\newcommand{\charv}{$\overline{char}$\xspace}
\newcommand{\logis}{$logical$\xspace}
\newcommand{\logiv}{$\overline{logical}$\xspace}
\newcommand{\dbls}{$double$\xspace}
\newcommand{\dblv}{$\overline{double}$\xspace}
\newcommand{\ints}{$integer$\xspace}
\newcommand{\intv}{$\overline{integer}$\xspace}

@@

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in \RLOC lines of R 
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by \PROGRAMMERS programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our result suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

\end{abstract}

@@

	\item if the analysis encounters \code{ typeof(x) == "double"}, it will look at the length of {\tt x} to determine if {\tt x} is a scalar or a vector, generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.
	This is also true for integer, complex, raw, logical, and character {\tt typeof}s (i.e., the primitive vectorized data types).
	A notable imperfection here is is that unit length vectors will appear as scalars, so later in our analysis we ensure that argument signatures containing both scalar/X and vector/X are collapsed down to contain only vector/X.

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
nret.translator <- function(dat, spss="out", ...) {
  ...
  if(identical(spss, "out")) {
    if(!is.vector(dat)) {
        stop(simpleError("...'dat' must be a vector!"))
      }
    ...
  } else {
    ...
    # here, dat is data.frame
    items.idx <- items.idx[order(names(dat[, items.idx]))]
  }
\end{lstlisting}}\caption{Example of a data.frame/(\dblv) argument (from {\tt klausR} package).}\label{fig:dfdbl}\end{figure}

@@

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
tee <- function (x, theta, D1, D2, phi)  {
    if (is.vector(x)) {
        x <- t(x)
    }
    ...
\end{lstlisting}}\caption{Example of matrix/(\dblv) argument (from {\tt calibrator} package).}\label{fig:matvec}\end{figure}

@@

\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \dbls & numeric & \{\} \\
y & \dblv & numeric & \{\} \\
retv & list<double> & list & \{\xspace example: \chars\}  
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  \dblv, \intv & 7742 & 9.66 \\ 
  \dbls, \ints & 7201 & 8.99 \\ 
  matrix, \dblv & 5384 & 6.72 \\ 
  NULL, \chars & 2325 & 2.90 \\ 
  list$<$any$>$, list$<$double$>$ & 2231 & 2.78 \\ 
  data.frame, matrix & 2116 & 2.64 \\ 
  NULL, \dbls & 2082 & 2.60 \\ 
  NULL, \dblv & 2069 & 2.58 \\ 
  list$<$any$>$, list$<$list$>$ & 1432 & 1.79 \\ 
  NULL, \charv & 1380 & 1.72 \\
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the L1.}
\end{table}

We would like to make one important note before digesting the data in Table~\ref{tab:toppolyL1}.
We drew a distinction between scalars and vectors of the same type, and that distinction would be the single largest source of polymorphism in L1 had we not dealt with it appropriately.
We note that an argument which exhibits a signature including e.g. \{\dbls, \dblv\} really must be an argument designed to take vectors, as recall that unit length vectors appear as scalars by our definition.
We have already collapsed this distinction.

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Tue Mar 26 17:46:45 2019
\begin{table}[ht]
\label{tab:toppolyL1p}
\centering
\begin{tabular}{lrr}
  \hline
Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  matrix, \realv & 6360 & 13.94 \\ 
  list$<$any$>$, list$<$real$>$ & 2799 & 6.13 \\ 
  \charv, \realv & 2774 & 6.08 \\ 
  \chars, \reals & 2552 & 5.59 \\ 
  data.frame, matrix & 2167 & 4.75 \\ 
  list$<$any$>$, list$<$list$>$ & 1496 & 3.28 \\ 
  list$<$any$>$, list$<$character$>$ & 1214 & 2.66 \\ 
  data.frame, \realv & 1046 & 2.29 \\ 
  \logis, \reals & 920 & 2.02 \\ 
  \chars, \realv & 868 & 1.90 \\
   \hline
\end{tabular}
\caption{Top 10 argument signatures in L$1^{+}$.}
\end{table}

A few of the top polymorphic signatures stand out.
First, the matrix/\realv signature is interesting: \AT{as we showed in our polymorphism examples section}, converting between vectors and matrices is very simple, and indeed numeric vectors are mathematically {\tt 1 x n} matrices.
In fact, in R matrices can be used identically to vectors in many cases, and we refer the reader to Figure~\ref{fig:matasvec} for examples.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
v <- c(1, 2, 3, 4)
ma <- matrix(data=v, nrow=4, ncol=1)
v[2] == m[2] == m[2, 1]
